\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-04-11 paper.qmd
math changes for the mappings

We define the NLDR as a function $g\text{:}~ \mathbb{R}^p \rightarrow \mathbb{R}^k$, with (hyper-)parameters $\mathbfit{\theta}$. The parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that maps the entire space $\mathbb{R}^p$ to a subset of $\mathbb{R}^k$ is suitable. <!--Any input requirements for the data (such as normalization, or preprocessing through the use of PCA or similar) is considered part of the function $g$.-->

@@

Define a hexagonal grid across the \kD{} space, using hexagons with fixed height ($h_b$), width ($w_b$). The starting coordinates $(x_s, y_s)$ mark the lower left of the grid. <!--We deliberately separate out the creation of the hexagonal grid from the mapping of points on the grid.-->
Each of the \kD{} points will belong to a hexagonal bin. That is, for each $y \in \mathbfit{Y}$, we can (uniquely) identify the hexagon that the point belongs to. Let $h : \mathbb{R}^k \rightarrow \mathbb{R}^k$ be the function that maps a point in \kD{} space to it's hexagonal centroid ($C^{(k)}$). <!--It follows that $h \circ g : \mathbb{R}^p \rightarrow \mathbb{R}^k$ maps each point in $\mathbfit{X}$ to a \kD{} centroid.-->

<!--First, a random variable $a$ is generated from a uniform distribution between $-\frac{\pi}{2}$ and $\frac{\pi}{2}$, representing the parameter $a$ is in $T$. Next, three variables $x_1$, $x_2$, and $x_3$ are computed based on $a$ to construct the S-curve data. $x_1$ is calculated as the sine of $a$ (@eq-eqn1), $x_2$ as twice the value of another uniformly distributed random variable between 0 and 1 (@eq-eqn2), and $x_3$ as the cosine of $a$ minus 1, multiplied by the sign of $a$ (@eq-eqn3). Additionally, four noise dimensions are introduced to the dataset to introduce noise variability: $x_4$, $x_5$, $x_6$, and $x_7$, each generated from uniform distributions within specific intervals (@eq-eqn4, @eq-eqn5, and @eq-eqn6 respectively). Finally, these variables are combined to form the S-curve dataset $T$.

$$
x_1 = \sin(3 \pi a)
$$ {#eq-eqn1}

$$
x_2 = 2.0 \times U(0, 1)
$$ {#eq-eqn2}

$$
x_3 = \text{sign}(a) \times (\cos(3 \pi a) - 1)
$$ {#eq-eqn3}

$$
x_4, x_5 \sim U(-0.02, 0.02)
$$ {#eq-eqn4}

$$
x_6 \sim U(-0.1, 0.1)
$$ {#eq-eqn5}

$$
x_7 \sim U(-0.01, 0.01)
$$ {#eq-eqn6}
-->

@@

```{r}
#| label: fig-NLDR-scurve
#| echo: false
#| fig-cap: "Key steps for constructing the $2-D$ model for S-curve data: (a) obtain UMAP data, (b) hexagonating UMAP data, (c) obtain hexagonal bin centroids, and (d) triangulate bin centroids."
#| fig-width: 8
#| fig-height: 4
#| out-width: 100%

>>>>

We define the NLDR as a function $g\text{:}~ \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{n\times k}$, with (hyper-)parameters $\mathbfit{\theta}$. The parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that does this mapping is suitable. <!--Any input requirements for the data (such as normalization, or preprocessing through the use of PCA or similar) is considered part of the function $g$.-->

@@

Define a hexagonal grid across the \kD{} space, using hexagons with fixed height ($h_b$), width ($w_b$). The starting coordinates $(x_s, y_s)$ mark the lower left of the grid. <!--We deliberately separate out the creation of the hexagonal grid from the mapping of points on the grid.-->
Each of the \kD{} points will belong to a hexagonal bin. That is, for each $y \in \mathbfit{Y}$, we can (uniquely) identify the hexagon that the point belongs to. Let $h : \mathbb{R}^{n\times k} \rightarrow \mathbb{R}^{m\times k}$ be the function that maps a point in \kD{} space to it's hexagonal centroid ($C^{(k)}$).

@@

```{r}
#| label: fig-NLDR-scurve
#| echo: false
#| fig-cap: "Key steps for constructing the model on the NLDR layout ($k=2$) of the S-curve data: (a) obtain UMAP data, (b) hexagonating UMAP data, (c) obtain hexagonal bin centroids, and (d) triangulate bin centroids."
#| fig-width: 8
#| fig-height: 4
#| out-width: 100%

### 2024-04-11 paper.tex
math changes for the mappings

We define the NLDR as a function
\(g\text{:}~ \mathbb{R}^p \rightarrow \mathbb{R}^k\), with
(hyper-)parameters \(\mathbfit{\theta}\). The parameters,
\(\mathbfit{\theta}\), depend on the choice of \(g\), and can be
considered part of model fitting in the traditional sense. Common
choices for \(g\) include functions used in tSNE, UMAP, PHATE, TriMAP,
PaCMAP, or MDS, although in theory any function that maps the entire
space \(\mathbb{R}^p\) to a subset of \(\mathbb{R}^k\) is suitable.

@@

Define a hexagonal grid across the \kD{} space, using hexagons with
fixed height (\(h_b\)), width (\(w_b\)). The starting coordinates
\((x_s, y_s)\) mark the lower left of the grid. Each of the \kD{} points
will belong to a hexagonal bin. That is, for each
\(y \in \mathbfit{Y}\), we can (uniquely) identify the hexagon that the
point belongs to. Let \(h : \mathbb{R}^k \rightarrow \mathbb{R}^k\) be
the function that maps a point in \kD{} space to it's hexagonal centroid
(\(C^{(k)}\)).

@@

\caption{\label{fig-NLDR-scurve}Key steps for constructing the \(2-D\)
model for S-curve data: (a) obtain UMAP data, (b) hexagonating UMAP
data, (c) obtain hexagonal bin centroids, and (d) triangulate bin
centroids.}

>>>>

We define the NLDR as a function
\(g\text{:}~ \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{n\times k}\),
with (hyper-)parameters \(\mathbfit{\theta}\). The parameters,
\(\mathbfit{\theta}\), depend on the choice of \(g\), and can be
considered part of model fitting in the traditional sense. Common
choices for \(g\) include functions used in tSNE, UMAP, PHATE, TriMAP,
PaCMAP, or MDS, although in theory any function that does this mapping
is suitable.

@@

Define a hexagonal grid across the \kD{} space, using hexagons with
fixed height (\(h_b\)), width (\(w_b\)). The starting coordinates
\((x_s, y_s)\) mark the lower left of the grid. Each of the \kD{} points
will belong to a hexagonal bin. That is, for each
\(y \in \mathbfit{Y}\), we can (uniquely) identify the hexagon that the
point belongs to. Let
\(h : \mathbb{R}^{n\times k} \rightarrow \mathbb{R}^{m\times k}\) be the
function that maps a point in \kD{} space to it's hexagonal centroid
(\(C^{(k)}\)).

@@

\caption{\label{fig-NLDR-scurve}Key steps for constructing the model on
the NLDR layout (\(k=2\)) of the S-curve data: (a) obtain UMAP data, (b)
hexagonating UMAP data, (c) obtain hexagonal bin centroids, and (d)
triangulate bin centroids.}
